---
title: "Metric for Batch Effects"
output: html.document
date: "2022-10-28"
---

```{r setup, include=FALSE}
library(MASS)
library(kBET)
library(ggplot2)
library(lisi)
```

We first set some hyper-parameters (mean and variance) for different cell types.

```{r}
cell.mu <- list(c(-3, 0), c(3, 0))
cell.sigma <- rep(list(matrix(c(1, 0, 0, 1), 2, 2)), 2)
```

And we also create random distributions for batch effects for different batches. The batch effects are set to be smaller than biological variance.

```{r}
# Distribution of batch effects
be.mu <- list(c(0, 1), c(1, -1))
be.sigma <- list(matrix(c(0.5, 0, 0, 0.5), 2, 2), matrix(c(0.5, 0, 0, 0.5), 2, 2))
```

Now, we can start creating data for batches.

```{r}
num.cells <- 1000
batch.composition <- list(c(0.7, 0.3), c(0.3, 0.7))
generate.data <- function(num.cells, batch.composition) {
  for (i in 1:length(batch.composition)) {
    batch.composition[[i]] = batch.composition[[i]] * num.cells
  }
  
  # data for different batch, without batch effect
  batches <- list()
  for (i in 1:length(batch.composition)) {
    batch <- data.frame()
    for (j in 1:length(cell.mu)) {
      cells <- mvrnorm(n = batch.composition[[i]][j], cell.mu[[j]], cell.sigma[[j]])
      cells <- data.frame(cells)
      cells$cell.type <- j
      batch <- rbind(batch, cells)
    }
    batch$batch <- as.character(i)
    batches <- append(batches, list(batch))
  }

  # data with batch effect
  batches.be <- list()
  for (i in 1:length(batch.composition)) {
    be <- mvrnorm(n = nrow(batches[[i]]), be.mu[[i]], be.sigma[[i]])
    batch.be <- batches[[i]][c("X1", "X2")] + be
    batch.be[c("cell.type", "batch")] <- batches[[i]][c("cell.type", "batch")]
    batches.be <- append(batches.be, list(batch.be))
  }
  return(list(batch=batches, batch.be=batches.be))
}
res <- generate.data(num.cells, batch.composition)
batches <- res$batch
batches.be <- res$batch.be
```


We can plot out the batches,

```{r}
# function for plotting batches out
plot_batches <- function(batch, title) {
  X1.min = min(batch[[1]]$X1)
  X1.max = max(batch[[1]]$X1)
  X2.min = min(batch[[1]]$X2)
  X2.max = max(batch[[1]]$X2)
  for (i in 2:length(batch)) {
    X1.min = min(X1.min, batch[[i]]$X1)
    X1.max = max(X1.min, batch[[i]]$X1)
    X2.min = min(X2.min, batch[[i]]$X2)
    X2.max = max(X2.max, batch[[i]]$X2)
  }

  
  
  plot(batch[[1]]$X2 ~ batch[[1]]$X1, 
       xlab = "Gene A", 
       ylab = "Gene B", 
       pch = batch[[1]]$cell.type, 
       col = 1, 
       xlim = c(X1.min, X1.max+3), 
       ylim = c(X2.min-1, X2.max+1),
       main = title)
  for (i in 2:length(batch)) {
    points(batch[[i]]$X2 ~ batch[[i]]$X1, 
           pch=batch[[i]]$cell.type, 
           col=i)
  }
  legend("bottomright", 
         title = "cell type", 
         legend = letters[1:length(cell.mu)], 
         pch = 1:length(cell.mu))
  legend("topright", 
         title = "batch", 
         legend = 1:length(be.mu), 
         pch = c(15, 15), 
         col = 1:length(be.mu))
}
plot_batches(batches, "Batches without batch effect")
```

```{r}
plot_batches(batches.be, "Batcees with batch effect")
```

Now we can try using kBET on the dataset,

```{r}
res <- generate.data(num.cells, list(c(0.1, 0.9), c(0.9, 0.1)))
batches <- res$batch
batches.be <- res$batches.be
data <- NULL
data.batch <- NULL
for (i in 1:length(batches)) {
  data <- rbind(data, batches[[i]][c("X1", "X2")])
  data.batch <- c(data.batch, batches[[i]]$batch)
}
kBET.res <- kBET(data, data.batch)
mean(kBET.res$stats$kBET.observed)
```

And for data with batch effect,

```{r}
LISI.res <- compute_lisi(data, data.frame(data.batch), "data.batch")
mean(LISI.res$data.batch)
```

Next, we try to fit projections of each batch on individual genes.

```{r}
library(mclust)
fit.gmm <- function(data, color, label) {
  fit = Mclust(data, modelName = "V")
  plot(fit, what = "density", xlim=c(-7, 7), col=color, xlab=label)
  rug(data, col=color)
  return(fit)
}
B1.X1.fit <- fit.gmm(batches.be[[1]]$X1, 1, "Batch 1 Gene A")
B2.X1.fit <- fit.gmm(batches.be[[2]]$X1, 2, "Batch 2 Gene A")
B1.X2.fit <- fit.gmm(batches.be[[1]]$X2, 1, "Batch 1 Gene B")
B2.X2.fit <- fit.gmm(batches.be[[2]]$X2, 2, "Batch 2 Gene B")
fit <- list(list(B1.X1.fit, B2.X1.fit), list(B1.X2.fit, B2.X2.fit))

```

We can do resampling based on the fitted GMM.

```{r}
lm.data <- NULL
for (i in 1:length(be.mu)) {
  resample.data <- NULL
  for (j in 1:length(B1.X1.fit$parameters$pro)) {
    resample <- rnorm(n = 1000 * fit[[1]][[i]]$parameters$pro[j], mean =  fit[[1]][[i]]$parameters$mean[j], sd = sqrt(fit[[1]][[i]]$parameters$variance$sigmasq[j]))
    resample <- data.frame(Y = resample, GC = as.character(j))
    resample.data <- rbind(resample.data, resample)
  }
  resample.data$batch <- as.character(i)
  hist(resample.data$Y)
  lm.data <- rbind(lm.data, resample.data)
}


```



```{r}
X1.lm <- lm(Y~GC + batch, lm.data)
summary(X1.lm)

```

```{r}
ggplot(lm.data, aes(x = batch, y = Y)) + geom_boxplot() 
```